\documentclass[11pt]{article}

% ------------------------------------------------------------------
% MiniChess AI -- Mini Project 2 report
% Compile: pdflatex report.tex  (uses only standard packages)
% ------------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}

\definecolor{codebg}{rgb}{0.96,0.96,0.96}
\definecolor{kw}{rgb}{0.0,0.0,0.6}
\definecolor{cmt}{rgb}{0.0,0.45,0.0}
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{codebg},
  keywordstyle=\color{kw}\bfseries,
  commentstyle=\color{cmt}\itshape,
  breaklines=true,
  frame=single,
  columns=fullflexible,
  showstringspaces=false,
  language=C++,
  numbers=none,
}
\newcommand{\code}[1]{\texttt{#1}}

\title{\textbf{MiniChess AI -- Design, Search, and a Neural-Network Evaluation Study}\\
\large Mini Project 2}
\author{Student ID: \texttt{114000110}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
This report documents a strong AI for the 6$\times$5 ``MiniChess'' variant and is
written to be self-contained: it explains not only \emph{what} the engine does
but \emph{why} each choice is correct, so every design decision can be defended
without referring to the source. The engine combines a classical
alpha-beta / Principal Variation Search (PVS) with a full modern search stack
(transposition table, killer/counter/history move ordering, iterative deepening,
aspiration windows, null-move pruning, late-move reductions, futility pruning,
quiescence search) and a \emph{hybrid evaluation} that blends a hand-crafted
heuristic with a small \emph{NNUE} (Efficiently Updatable Neural Network) trained
from scratch on self-play data. Beyond the working engine, the report presents an
empirical study: porting Stockfish's NNUE architecture ideas (SFNNv1--v14), int16
quantization, texel tuning, and a search-allocation optimization. The central,
repeatedly-confirmed finding is that \textbf{on a tiny board, search depth
dominates evaluation sophistication}, and that \textbf{training-loss improvements
do not imply playing-strength improvements}. The final engine beats every
provided baseline (weak/strong minimax, alpha-beta, PVS) with both colours and
sweeps the visible boss 8/8. Section~\ref{sec:qa} is a TA-question crib sheet.
\end{abstract}

\tableofcontents

% ==================================================================
\section{Overview and architecture}
% ==================================================================
The program speaks \textbf{UBGI} (a UCI-superset protocol) over stdin/stdout.
The code is split into game-agnostic and game-specific layers:

\begin{itemize}[nosep]
  \item \code{src/ubgi/ubgi.cpp} -- protocol loop and the \emph{iterative
        deepening} driver (calls the search at depth $1,2,3,\dots$, honours
        \code{go movetime}, emits \code{info}/\code{bestmove}).
  \item \code{src/games/minichess/state.cpp} -- board representation, move
        generation (naive + bitboard), Zobrist hashing, terminal detection, the
        reference hand-crafted evaluation.
  \item \code{src/policy/submission.cpp} -- the engine we developed: PVS search,
        all pruning/ordering heuristics, the tuned hand-crafted evaluation, and
        the NNUE blend. (This is the policy that competes; algorithm name
        \code{submission}.)
  \item \code{src/policy/minimax.cpp} -- a clean reference PVS+quiescence used to
        explain the core algorithm and to generate early training games.
  \item \code{src/policy/nnue.hpp}, \code{nnue\_weights.hpp} -- the neural
        evaluator and its embedded weights (plain C++ arrays).
  \item \code{tools/nnue\_gen.cpp} -- a C++ self-play data generator that labels
        positions with the strong \code{submission} search; \code{tools/train\_nnue.py}
        -- a PyTorch (GPU) trainer that consumes the generated data and emits the
        embedded weights. (Training is offline only; the engine itself uses no
        external ML dependencies.)
\end{itemize}

\paragraph{The big picture.} Each turn the driver runs iterative deepening; at
every node the search applies alpha-beta/PVS with aggressive but provably-safe
pruning; at the leaves a quiescence search resolves captures and then calls the
hybrid evaluation (hand-crafted $+$ NNUE). Move generation, hashing, and the
evaluation are all expressed from the \emph{side-to-move's} perspective so the
search is a single symmetric negamax.

% ==================================================================
\section{Game rules, board representation, and move generation}\label{sec:rules}
% ==================================================================
\subsection{Rules (and the one rule that shapes everything)}
MiniChess is chess on a 6-row $\times$ 5-column board with a \textbf{king-capture}
win model: \emph{a side wins the instant it can capture the enemy king on its
turn}. There is no castling and no en passant; a pawn moves one square forward
(no double step) and promotes \emph{only to a queen} on the last rank. Draws occur
by 3-fold repetition and by a step cap; if the cap is hit the winner is decided by
summed piece value.

The king-capture model is the single most important simplification, and it
drives several design choices (Section~\ref{sec:qa}):
\begin{itemize}[nosep]
  \item \textbf{There is no ``illegal move''.} You are \emph{allowed} to move into
        a square where your king can be captured next turn -- you simply lose then.
        Therefore every move produced by the generator is legal, and \textbf{the
        engine can never lose by outputting an illegal move.}
  \item \textbf{No checkmate/stalemate detection is needed.} A terminal ``win''
        is detected purely as ``a generated move lands on the enemy king.''
        Losing is handled implicitly: if we leave our king capturable, the
        opponent's reply node is a win \emph{for them}, which negamax turns into a
        loss for us.
\end{itemize}

\subsection{Board representation and move encoding}
The board is two planes, one per player:
\begin{lstlisting}
char board[2][6][5];  // board[player][row][col]; values:
// 0 empty, 1 Pawn, 2 Rook, 3 Knight, 4 Bishop, 5 Queen, 6 King
\end{lstlisting}
A \code{Move} is \code{pair<Point,Point>} $=$ (from, to) with \code{Point =
pair<size\_t,size\_t>} $=$ (row, col). Promotion is implicit: a pawn reaching row
$0$ (White) or row $5$ (Black) is rewritten to a queen inside \code{next\_state}.
Row~$0$ is the top; White starts on rows 4--5 and advances toward row~0.

\subsection{Move generation: bitboards on 30 bits}
$6\times5 = 30$ squares fit in a \code{uint32\_t}, square $(r,c)\to$ bit $r{\cdot}5{+}c$.
At startup we precompute leaper attack masks (knight, king, pawn-push, pawn-capture).
Generation builds occupancy masks, then bit-scans own pieces with
\code{\_\_builtin\_ctz}; leapers index their mask, sliders (rook/bishop/queen)
walk direction rays until blocked. The naive array version
(\code{get\_legal\_actions\_naive}) is kept for reference; the engine uses the
bitboard path (\code{USE\_BITBOARD}).

\begin{lstlisting}[caption={Sliding-piece generation + early king-capture win (bitboard)}]
int d_start = (piece==4)?4:0, d_end = (piece==2)?4:8; // R:0-3 B:4-7 Q:0-7
for(int d=d_start; d<d_end; d++){
    int cr=r+bb_dr[d], cc=c+bb_dc[d];
    while(in_bounds(cr,cc)){
        uint32_t to_bit = 1u << BB_SQ(cr,cc);
        if(self_occ & to_bit) break;             // own piece blocks
        if((oppn_occ & to_bit) && oppn_pt[to]==6){// captures enemy KING:
            game_state = WIN;                     // -> we win right now
            legal_actions.push_back(Move({r,c},{cr,cc}));
            return;                               // short-circuit: nothing else matters
        }
        targets |= to_bit;
        if(oppn_occ & to_bit) break;              // capture stops the ray
        cr += bb_dr[d]; cc += bb_dc[d];
    }
}
\end{lstlisting}
The \code{game\_state=WIN; return;} pattern is the terminal detector: the moment
any piece can take the enemy king, generation stops and flags a win, so the
search treats the node as a won leaf.

\subsection{State transition and the node pool}
\code{next\_state(move)} copies the board, applies the move (with promotion),
incrementally updates the Zobrist hash, flips the side to move, and returns a new
\code{State}. For speed the search instead uses \code{apply\_into(dst,move)} /
\code{null\_into(dst)}, which write the child position \emph{in place} into a
caller-owned pool slot, avoiding a heap allocation per node
(Section~\ref{sec:nodepool}).

% ==================================================================
\section{Position keys: Zobrist hashing and repetition}\label{sec:zobrist}
% ==================================================================
Each (piece, colour, square) and the side-to-move own a random 64-bit key
(xorshift-seeded once). A position's hash is the XOR of the keys of all occupied
squares, plus the side key if Black is to move. The hash is updated
\emph{incrementally} on each move -- XOR out the moving piece at its source, XOR
out any captured piece, XOR in the (possibly promoted) piece at the destination,
and toggle the side key:
\begin{lstlisting}[caption={Incremental Zobrist update in next\_state}]
uint64_t h = this->hash();
h ^= zobrist_side;                                   // side to move flips
h ^= zobrist_piece[p][orig_piece][from.r][from.c];   // remove from source
if(captured) h ^= zobrist_piece[opp][captured][to.r][to.c]; // remove victim
h ^= zobrist_piece[p][moved][to.r][to.c];            // add at destination (moved may be promoted Q)
\end{lstlisting}
This makes the hash $O(1)$ per move instead of $O(30)$. The hash keys the
transposition table (Section~\ref{sec:tt}) and the repetition map.

\paragraph{Repetition / draws.} \code{GameHistory} is a \code{hash $\to$ count}
map; the search pushes the current hash before recursing and pops it after.
\code{check\_repetition} returns a draw ($0$) when a position has occurred
\textbf{3 times}, implementing the 3-fold rule and, crucially, stopping the search
from looping forever in a repetition.

% ==================================================================
\section{State value function (evaluation)}
% ==================================================================
The evaluation returns an integer in centipawn-like units \emph{from the
side-to-move's perspective} (so the search is a pure negamax). The shipped engine
(\code{submission.cpp}) uses a \textbf{blend} of a tuned hand-crafted term and the
NNUE. (The reference \code{minimax.cpp}/\code{state.cpp} use a simpler
hand-crafted eval -- material$+$PST$+$tropism$+$mobility -- which is useful to
contrast at the demo but is not what competes.)

\subsection{Hand-crafted evaluation (shipped, \texttt{custom\_evaluate})}
For each side we sum, then take self $-$ opponent:
\begin{equation}
S_{\text{hand}} = (\text{material}+\text{PST}+\text{tropism}+\text{pawns})_{\text{self}}
  - (\cdots)_{\text{opp}} + w_m\cdot|\text{legal moves}|,
\end{equation}
with these components:
\begin{itemize}[nosep]
  \item \textbf{Material} $\{P,R,N,B,Q,K\}=\{100,500,320,330,900,10000\}$. The
        king's $10000$ is a sentinel; real king safety comes from search, not this
        number.
  \item \textbf{Piece-square tables (PST)}: a per-piece $6\times5$ table, read from
        the side-to-move's orientation (Black's row is mirrored, \code{pst\_r =
        player?\,r:\,5-r}). Knights/bishops/queens are pushed toward the centre;
        the king is pushed back (negative centre, positive home squares).
  \item \textbf{King tropism}: own non-king pieces get
        $w_t[\text{type}]\cdot\max(0,4-\text{Chebyshev dist to enemy king})$ with
        weights $\{P,R,N,B,Q\}=\{2,5,8,5,10\}$ -- rewards crowding the enemy king,
        which directly ends the game.
  \item \textbf{Pawn structure}: doubled pawns $-15$ each extra on a file;
        isolated pawns (no friendly pawn on an adjacent file) $-10$.
  \item \textbf{Passed pawns}: $\max(10, 50-10\cdot\text{dist-to-promotion})$ --
        large because promotion to a queen is decisive on a small board.
  \item \textbf{Mobility}: $w_m=5$ times the side-to-move's legal-move count
        (re-using the already-generated \code{legal\_actions}, so it is free).
\end{itemize}
Terminal positions short-circuit before any of this: a win returns $P_{\max}$, a
draw returns $0$ (and the search adjusts wins to $P_{\max}-\text{ply}$, below).

\subsection{NNUE evaluation and the blend}\label{sec:nnueeval}
The NNUE (Section~\ref{sec:nnue}) produces a score from the same side-to-move
perspective; it is blended with the hand-crafted score:
\begin{equation}
S = \frac{b\cdot S_{\text{nnue}} + (100-b)\cdot S_{\text{hand}}}{100},
\qquad b = \texttt{NNUE\_BLEND} = 45.
\end{equation}
NNUE is on by default and self-contained; \code{-DNO\_NNUE} disables it and
\code{-DNNUE\_BLEND=$n$} overrides the mix. \textbf{Why a blend, and why 45\%?}
Empirically (Section~\ref{sec:blend}) pure NNUE is \emph{weaker} than the
hand-crafted eval on this game, but a partial admixture is meaningfully stronger
than either component alone: the net captures outcome-correlated patterns the
heuristic misses, while the heuristic supplies tactical robustness and is cheap.

% ==================================================================
\section{Tree search}
% ==================================================================
\subsection{From minimax to negamax}
Minimax assumes the opponent minimises our score. Because MiniChess strictly
alternates turns and the evaluation is symmetric (side-to-move relative), we use
the \emph{negamax} identity $\text{value}(n)=\max_m -\text{value}(\text{child})$:
each side maximises its own perspective, and a child's score is negated when read
by the parent.

\begin{lstlisting}[caption={Negamax with alpha-beta (concept)}]
int negamax(node, depth, alpha, beta):
    if terminal(node): return WIN? P_MAX-ply : (DRAW? 0 : ...)   // mate-distance aware
    if depth == 0: return quiescence(node, alpha, beta)
    for each move m (in ordering priority):
        score = -negamax(child(m), depth-1, -beta, -alpha)
        if score >= beta: return beta            // fail-high: opponent avoids this line
        if score > alpha: alpha = score          // new best
    return alpha
\end{lstlisting}

\paragraph{Mate scoring.} A win returns $P_{\max}-\text{ply}$ (not just
$P_{\max}$). Subtracting the ply makes \emph{shorter} mates score higher, so the
engine prefers to win quickly and to delay losing -- and it lets mate scores be
compared meaningfully across the tree. TT storage adds the ply back on store and
removes it on probe so a mate found via a transposition keeps the correct
distance.

\paragraph{King-capture and the ``same player'' flag.} The generic search
supports games where the mover does not alternate (e.g.\ Connect6) via
\code{same\_player\_as\_parent()}; MiniChess always alternates, so the child is
always negated. Because a capturable enemy king makes \code{get\_legal\_actions}
flag \code{WIN}, the negamax terminal check fully captures both winning (our move
takes their king) and losing (their reply takes ours, negated) without any
special-casing.

\subsection{Alpha-beta pruning}
$\alpha$ is the best score the side to move is already assured of; $\beta$ is the
best the opponent is assured of. When $\alpha\ge\beta$ the remaining moves cannot
affect the result (the opponent would never enter this line), so the branch is
pruned. Alpha-beta returns the \emph{exact same} value as minimax but, with good
ordering, visits $O(b^{d/2})$ instead of $O(b^d)$ nodes -- effectively doubling
reachable depth.

\subsection{Principal Variation Search (PVS)}
PVS assumes the first (best-ordered) move is best. It searches that move with the
full window $[\alpha,\beta]$ and every later move with a \emph{null window}
$[\alpha,\alpha{+}1]$ -- a cheap boolean test ``is this move worse than the best
so far?''. If a null-window search surprises us by failing high, that move is
re-searched with the full window. When ordering is good, almost all moves are
refuted by the cheap test, so PVS is strictly faster than plain alpha-beta with no
loss of correctness.

\begin{lstlisting}[caption={PVS core (from submission.cpp, simplified)}]
if(move_index == 0){
    score = -search(child, depth-1, -beta, -alpha);       // full window on PV move
}else{
    score = -search(child, depth-1, -alpha-1, -alpha);    // null-window scout
    if(score > alpha && score < beta)                     // scout failed high
        score = -search(child, depth-1, -beta, -alpha);   // re-search full window
}
\end{lstlisting}

\subsection{Quiescence search}
At depth~0 the engine does not evaluate immediately; it runs a \emph{quiescence}
search over ``noisy'' moves (captures and promotions) until the position is quiet,
avoiding the \emph{horizon effect} (evaluating in the middle of a capture
sequence). It uses (i)~a \textbf{stand-pat} cutoff -- the static eval is a lower
bound because the side to move can usually decline to capture; (ii)~MVV-LVA
ordering; and (iii)~\textbf{delta pruning} -- skip a capture if
$\text{stand-pat}+V_{\text{victim}}+\text{margin}<\alpha$ (it cannot raise
$\alpha$ even in the best case).

\begin{lstlisting}[caption={Quiescence stand-pat and delta pruning}]
int stand_pat = custom_evaluate(state);
if(stand_pat >= beta) return beta;                    // already good enough
if(stand_pat + (QUEEN+200) < alpha) return alpha;     // whole-node delta cutoff
if(alpha < stand_pat) alpha = stand_pat;
// ... then try only captures/promotions, MVV-LVA ordered ...
\end{lstlisting}

\subsection{Transposition table (TT)}\label{sec:tt}
A $2^{22}\approx4{\cdot}10^6$-entry hash table caches search results keyed by the
Zobrist hash. Each \code{TTEntry} ($\approx$10\,bytes; $\approx$40\,MB total)
stores a 32-bit key check (to detect index collisions), score, depth, a bound flag
(\textbf{exact}/\textbf{lower}/\textbf{upper}), a 16-bit packed best move
(\code{CompactMove}, 4 bits per coordinate), and a generation byte for ageing. The
TT (i)~returns a usable score when the position recurs at $\ge$ the stored depth
\emph{and} the bound is compatible with the current window, and (ii)~feeds the
\emph{TT move} to move ordering -- the single biggest ordering win.

\begin{lstlisting}[caption={Bound-aware TT probe (why a cached score is usable)}]
if(entry.depth >= depth){
    if(entry.flag == TT_EXACT)              return score;          // exact value
    if(entry.flag == TT_ALPHA && score<=alpha) return alpha;       // upper bound <= alpha
    if(entry.flag == TT_BETA  && score>=beta)  return beta;        // lower bound >= beta
}
\end{lstlisting}
An \emph{exact} entry is the true value; a \emph{lower} bound (fail-high) is only
trusted when it already exceeds $\beta$; an \emph{upper} bound (fail-low) only when
it is $\le\alpha$. This is exactly the information alpha-beta proved when the entry
was stored, so reusing it changes nothing about correctness.

\subsection{Move ordering}
Good ordering is what makes alpha-beta/PVS effective. Moves are tried in this
priority (higher score first):
\begin{enumerate}[nosep]
  \item \textbf{TT move} -- the best move previously found here ($\approx 10^7$).
  \item \textbf{Captures} by \textbf{MVV-LVA} (Most Valuable Victim, Least Valuable
        Attacker: $16\,V_{\text{victim}}-V_{\text{attacker}}$).
  \item \textbf{Promotions}.
  \item \textbf{Killer moves}: two quiet moves per ply that recently caused a
        cutoff at the same ply.
  \item \textbf{Counter move}: the quiet move that previously refuted the
        opponent's last move (indexed by that move's from/to).
  \item \textbf{History heuristic}: quiet moves scored by how often they caused
        cutoffs, accumulated per (side, from, to) as $+d^2$, halved on overflow.
\end{enumerate}
Killers/counters/history are pure \emph{ordering} hints; they never prune, so a
bad hint costs a little speed but cannot change the result.

\subsection{Selectivity: pruning, reductions, extensions}
On top of exact alpha-beta the engine searches \emph{selectively}. Every technique
below is disabled in the situations where it would be unsound:
\begin{itemize}[nosep]
  \item \textbf{Null-move pruning (NMP)}: give the opponent a free move; if the
        position is still $\ge\beta$ at reduced depth ($R=\min(d{-}1,3{+}[d\!\ge\!7])$,
        clamped $\ge 2$), prune. Disabled in check and at PV nodes; unproven mate
        scores from a null search are clamped to $\beta$ (never returned as a real
        mate). Rationale: passing is almost always worse than the best real move,
        so if even a free move for the opponent can't drag us below $\beta$, no
        real move will.
  \item \textbf{Late-move reductions (LMR)}: late, quiet, non-killer moves are
        searched at reduced depth (reduction $1$--$4$, growing with move index and
        depth, $-1$ at PV nodes). If a reduced search beats $\alpha$ it is
        re-searched at full depth, so a reduction can only \emph{save time}, never
        drop a move that was actually good.
  \item \textbf{Reverse futility / static null move}: at low depth, if
        $\text{static}-120\,d\ge\beta$, return early (the position is so good a
        full search is unnecessary). Off in check / PV.
  \item \textbf{Futility pruning}: at low depth, skip quiet moves when
        $\text{static}+150\,d\le\alpha$ (they cannot plausibly raise $\alpha$).
  \item \textbf{Late-move pruning (LMP)}: at very low depth, skip quiet moves
        beyond a move-count threshold.
  \item \textbf{Mate-distance pruning}: tighten the window with known mate bounds
        so the engine prefers faster mates and never wastes effort proving a
        slower one.
  \item \textbf{Check extension}: search one ply \emph{deeper} when in check, so
        forcing lines are not cut off by the depth limit.
  \item \textbf{Internal iterative deepening (IID)}: at a PV node with no TT move,
        do a shallow search first to obtain a good first move for ordering.
\end{itemize}

\subsection{Iterative deepening, aspiration windows, time management}
The driver searches depth $1,2,\dots$, reusing the previous depth's TT and PV for
ordering (so a deeper search is usually \emph{faster} per node than starting
cold). From the previous score it opens a narrow \emph{aspiration window}
($\pm 50$); on a fail-high/low it widens the window exponentially and re-searches.
Time is checked every 2048 nodes; the search aborts shortly before the per-move
budget using a margin of $\min(300,\text{movetime}/25)$\,ms, and a ``half-time''
rule avoids starting a depth that almost certainly cannot finish. \textbf{The
competition limit is a strict 2\,s per move}; the engine adapts to whatever
\code{movetime} the runner passes and always has a legal move ready
(Section~\ref{sec:qa}, ``losing on time'').

% ==================================================================
\section{NNUE: a neural evaluation trained from scratch}\label{sec:nnue}
% ==================================================================
No pre-trained weights or runtime ML libraries are used; the engine embeds the
weights as C++ arrays. Training is offline (PyTorch on GPU) and feeds data
generated by a C++ self-play tool.

\subsection{Feature encoding (side-to-move relative)}
The board is canonicalised to the side to move so one set of weights serves both
colours. White keeps orientation; \textbf{Black is rotated $180^\circ$}
($r\!\to\!5{-}r,\ c\!\to\!4{-}c$), which is exact because Black's start array is
White's rotated $180^\circ$. Each occupied square contributes one binary feature:
\begin{equation}
\text{feature} = \underbrace{\text{base}}_{0\ \text{own},\ 180\ \text{enemy}}
 + (\text{piece}-1)\cdot 30 + \text{sq},\quad
\text{sq}=\begin{cases} r\cdot5+c & \text{White}\\ (5{-}r)\cdot5+(4{-}c) & \text{Black}\end{cases}
\end{equation}
giving $2\times6\times30 = 360$ features (own/enemy $\times$ 6 piece types $\times$
30 squares). \emph{Worked example}: a White rook on $(r{=}5,c{=}0)$ is own piece,
type 2, sq $=25$, so feature $=(2-1)\cdot30+25=55$. The encoding is implemented
\emph{identically} in \code{nnue\_gen.cpp} and \code{nnue.hpp} -- a mismatch would
silently corrupt the evaluation, so this is the first thing to check if the net
ever ``goes crazy.''

\subsection{Network and why it is small}
The shipped network is a fully-connected $360 \to 32 \to 1$ with ReLU and a
$\tanh$ output (scaled by $600$ to centipawns). The width is a deliberate
\emph{speed} decision: because the accumulator is rebuilt at every search leaf
(below), wider nets cost search depth. Measured at $1.5$\,s from the start
position:
\begin{center}\small
\begin{tabular}{@{}lcccc@{}}
\toprule
Hidden width $H$ & 32 & 64 & 128 & 256 \\
\midrule
Depth reached & \textbf{15} & 13 & 11 & 11 \\
\bottomrule
\end{tabular}
\end{center}
$H{=}128$ fits the data better (lower validation MSE) yet, in head-to-head play,
was \emph{no stronger} and sometimes worse -- the $\approx4$ lost plies cancel the
better eval. $H{=}32$ is the strength/speed optimum, so we kept it and improved
the \emph{weights} instead.

\subsection{Training pipeline}
\begin{enumerate}[nosep]
  \item \textbf{Self-play data} (\code{tools/nnue\_gen.cpp}): $\approx 30\text{k}$
        games are played by the strong \code{submission} search (depth-7 iterative
        deepening; random openings of $1$--$8$ plies plus $10\%$ random moves for
        diversity), yielding $\approx 926\text{k}$ labelled positions. Six
        processes generate shards in parallel; each non-terminal position stores
        its active features, side to move, the teacher search score, and (after
        the game) the result.
  \item \textbf{Labels}: each position carries its eventual game result (WDL,
        $\pm1/0$ from its side-to-move's perspective) \emph{and} the teacher search
        score. The target blends them,
        $t=\lambda\,\text{WDL}+(1-\lambda)\tanh(\text{score}/400)$ with
        $\lambda{=}0.7$ -- WDL keeps the net honest about outcomes, the teacher
        score sharpens positional nuance.
  \item \textbf{Training} (\code{tools/train\_nnue.py}, PyTorch on GPU): Adam,
        mini-batch MSE against the $\tanh$ output, cosine-decayed learning rate,
        $80$ epochs, $4\%$ validation hold-out (best-val checkpoint kept; final
        validation MSE $\approx 0.36$).
  \item \textbf{Embedding}: weights are written to \code{nnue\_weights.hpp} as
        plain C++ arrays (with $W_1$ laid out feature-major so the inner loop is
        contiguous), so the engine is self-contained and compiles with the
        standard command -- \textbf{no third-party library is linked into the
        engine}.
\end{enumerate}

\subsection{Inference: sparse, per-leaf, float}
The board has $\le 20$ pieces, so the accumulator is rebuilt \emph{sparsely from
scratch} at every leaf -- summing only the $\le 20$ active feature rows of $W_1$ --
rather than maintained incrementally. This trades a little speed for zero
make/unmake accumulator bugs (a classic NNUE pitfall), and on this board the eval
is still cheap. The output is
\begin{equation}
\text{cp} = \mathrm{round}\Big(\tanh\big(b_2 + \textstyle\sum_h W_2[h]\cdot\mathrm{relu}(b_1[h]+\!\!\sum_{f\,\text{active}}\!\!W_1[f][h])\big)\cdot 600\Big).
\end{equation}
\begin{lstlisting}[caption={NNUE forward pass (nnue.hpp, simplified)}]
float acc[H];
for(int h=0; h<H; h++) acc[h] = B1[h];               // bias
for(each occupied square) {                          // <= 20 iterations
    int f = feature_index(piece, r, c, stm);         // own/enemy split, 180-flip for black
    const float* w = &W1[f*H];
    for(int h=0; h<H; h++) acc[h] += w[h];            // add this feature's row
}
float pre = B2;
for(int h=0; h<H; h++) pre += W2[h] * relu(acc[h]);
return lround(tanh(pre) * 600);                       // centipawns, stm perspective
\end{lstlisting}

\subsection{Blend tuning}\label{sec:blend}
We measure blend ratios with \emph{diverse openings} (Section~\ref{sec:method}).
Pure NNUE is weaker than the hand-crafted eval here; a partial admixture is best.
For the retrained net we swept $b\in\{30,45,60\}$ against a fixed reference engine:
$b{=}45$ scored best (and never worse), so it is the shipped default. Against the
previous $b{=}30$ net (same $360\!\to\!32\!\to\!1$ architecture, $\approx
98\text{k}$ training positions) the retrained $b{=}45$ engine scores
\textbf{60\%} over $30$ games with mild openings, at \emph{identical} search speed.
It sweeps the visible boss \textbf{8/8} both from the fixed start and with diverse
openings at 2\,s.

% ==================================================================
\section{Architecture study: porting Stockfish NNUE (SFNNv1--v14)}
% ==================================================================
We investigated whether Stockfish's NNUE architecture evolution transfers to
MiniChess. This is a likely demo discussion point and a deliberate set of
\emph{negative} results.

\subsection{What changed across SFNN versions}
\begin{table}[h]\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
Version & Key architectural idea & Transfers to MiniChess? \\
\midrule
SFNNv1 & HalfKP features; \emph{two-perspective} accumulator; & dual-persp / clipped-ReLU: yes; \\
       & clipped ReLU; deeper head; int quantization & HalfKP: too many features \\
SFNNv4 & HalfKAv2 (king included as a piece) & partial (we include the king) \\
SFNNv5 & HalfKAv2\_hm (horizontal mirroring) & no -- board is not L/R symmetric \\
SFNNv6 & L1 width doubled; \emph{piece-count output buckets} & buckets: cheap, plausible \\
v7--v14& L1 scaling $1024\!\to\!3072$, more data & scaling only \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Dual-perspective accumulator (SFNNv1 core)}
We implemented the defining NNUE idea: encode the position from \emph{both} the
side-to-move's and the opponent's viewpoint through a \emph{shared} feature
transformer, concatenate the two accumulators, and pass through a deeper head
($360 \to 64\,[\times2] \to 32 \to 1$) with clipped ReLU. Trained on pure WDL it
reached a much lower MSE ($0.237$ vs $0.56$) -- yet it \textbf{played worse}: the
two accumulators plus the bigger head cut NPS $\sim$3$\times$, costing
$\sim$1.5--2 plies of depth. The lower loss did not translate into strength.

\subsection{int16 quantization (auto-vectorized)}
To recover the lost speed we quantized to int16 (SFNNv1's speed trick): weights
scaled by $Q_A{=}512$ (clipped-ReLU unit) and $Q_W{=}256$ (layer weights), with
\emph{int32} accumulation laid out contiguously so the compiler auto-vectorizes
under \code{-march=native} (no hand-written SIMD; float-vs-int parity
$\approx$2.5\,cp). This recovered much of the speed but the dual-perspective net
was \emph{still} below the simple single-perspective net. Quantizing the
single-perspective net gave \emph{no} speedup (its float eval is already cheap and
int16 widening cancels the lane-count gain).

\subsection{Conclusion of the study}
\textbf{Stockfish's NNUE architecture is sound but does not transfer to
MiniChess.} On full chess, NNUE's eval cost is amortised by enormous search trees,
the hand-crafted alternative is far weaker, and int8 SIMD + incremental updates
keep NNUE cheap. On a $6\times5$ board none of that holds: a cheap eval plus more
search depth wins, so the structural sophistication (and even the quantization
speed trick) is a net loss. The shipped engine therefore uses the
\emph{single-perspective float} net.

% ==================================================================
\section{Search-speed optimization: a ply-indexed node pool}\label{sec:nodepool}
% ==================================================================
Profiling pointed at per-node memory churn: the search did \code{new
State}+\code{delete} (and a fresh move-list allocation) for every node. We added
\code{State::apply\_into} / \code{null\_into}, which write the child position
\emph{in place} into a caller-owned \code{State} taken from a ply-indexed pool
(\code{g\_node\_pool[ply]}), reusing each slot's move-list buffer. Depth-first
search guarantees \code{pool[child\_ply]} is free to overwrite once a subtree
returns, and a parent never reads a child slot while recursing -- so the search
tree is \emph{provably identical} (verified: same node counts and scores) while
two heap allocations per node disappear. This is the one optimization (of three
attempted) that shipped, because a speedup must be a provable non-regression
before it is trusted.

% ==================================================================
\section{Negative results (also useful at the demo)}
% ==================================================================
\begin{itemize}[nosep]
  \item \textbf{Wider NNUE} ($H{=}64/128/256$): lower validation loss, no strength
        gain -- the lost search depth cancels it. Kept $H{=}32$.
  \item \textbf{Texel-tuning the material values} to fit WDL improved the static
        loss by only $0.4\%$ and \emph{lost} head-to-head: static-eval loss is a
        poor proxy for strength, and the material values also drive move ordering
        and delta pruning, so reweighting hurt search efficiency.
  \item \textbf{Opening book}: with a strict \emph{per-move} 2\,s limit (no game
        clock) a book cannot bank time and the engine already plays the fixed
        opening strongly, so a book would only echo the engine's own move. Not
        shipped (it would be a no-op).
  \item \textbf{Pure search-score teacher for NNUE}: reached a very low MSE but
        regressed in play -- the net merely learned to imitate the existing eval
        rather than predict outcomes. (We kept a \emph{blend} of WDL and teacher
        score instead.)
\end{itemize}

% ==================================================================
\section{Results and methodology}\label{sec:method}
% ==================================================================
\paragraph{Measuring strength correctly.} Both engines are deterministic, so games
from the fixed start position are effectively one game repeated -- a ``20-game
match'' from the start cannot resolve small differences. We therefore use
\emph{random openings} (\code{--opening-plies N --seed S}) in the match runner;
all comparative numbers use diverse openings, and matches are run sequentially
(time-based results need an unloaded CPU).

\begin{table}[h]\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
Opponent & Result (final engine) & Notes \\
\midrule
Weak MiniMax (baseline)    & win, both colours & advances baseline \\
Strong MiniMax (baseline)  & win, both colours & advances baseline \\
Alpha-Beta (baseline)      & win, both colours & weaker than the PVS boss \\
PVS (visible boss)         & \textbf{8/8}      & 2\,s, fixed start \\
PVS boss, diverse openings & \textbf{8/8}      & 2\,s, random openings (harder) \\
Previous NNUE ($b{=}30$)   & \textbf{60\%}     & 30 games, mild openings, equal speed \\
\bottomrule
\end{tabular}
\caption{Final engine ($H{=}32$ retrained, blend 45, node pool) vs.\ opponents.}
\end{table}

\paragraph{Constraints respected.} Standard library only (NNUE weights are
embedded C++ arrays; PyTorch is used \emph{offline} for training and is never
linked into the engine); no manual SIMD/inline-asm (only compiler
auto-vectorization via \code{-march=native}); the engine's policy code is
single-threaded; TT $\approx$40\,MB, well under the 4\,GB limit.

% ==================================================================
\section{Complexity and memory summary}
% ==================================================================
\begin{table}[h]\centering\small
\begin{tabular}{@{}lll@{}}
\toprule
Component & Cost & Notes \\
\midrule
Move generation       & $O(30)$ bit-scan + ray walks & bitboard, leaper masks precomputed \\
\code{next\_state}/hash & $O(1)$ amortised & incremental Zobrist; in-place into pool \\
Search (ordered AB/PVS) & $\approx O(b^{d/2})$ & $b\approx$ 10--20 here \\
Transposition table   & $2^{22}$ entries, $\approx$40\,MB & 32-bit key check + ageing \\
NNUE inference        & $\le 20$ feature rows $\times\,32$ + $32$ & rebuilt per leaf, float \\
History/killers/counters & $O(1)$ updates & ordering only, never prune \\
\bottomrule
\end{tabular}
\end{table}

% ==================================================================
\section{Anticipated TA questions and answers}\label{sec:qa}
% ==================================================================
\begin{description}[leftmargin=1.2em,style=nextline]
\item[Q. Why negamax instead of separate min/max code?] The evaluation is
  side-to-move relative and turns strictly alternate, so $\max$ for one side is
  $\min$ for the other. Negamax expresses both with one routine by negating the
  child's score; less code, fewer bugs.
\item[Q. Does alpha-beta change the move minimax would pick?] No. Alpha-beta only
  prunes branches that provably cannot affect the value at the root (a side would
  never choose to enter them). It returns the identical minimax value; it is
  purely a speed optimization.
\item[Q. How do you know the engine never plays an illegal move (instant loss)?]
  The move generator only ever produces legal moves, and in this king-capture
  variant there is no ``moving into check is illegal'' rule -- every generated
  move is legal. The engine always returns a move from \code{legal\_actions}, so
  an illegal move is structurally impossible.
\item[Q. How is winning/losing detected without checkmate logic?] Generation
  flags \code{WIN} the instant a move can capture the enemy king, and the search
  treats that node as a won leaf ($P_{\max}-\text{ply}$). Losing needs no separate
  code: if we leave our king capturable, the opponent's reply node is a win for
  them, which negamax negates into a loss for us.
\item[Q. Why $P_{\max}-\text{ply}$ rather than a flat $P_{\max}$?] So the engine
  prefers the \emph{fastest} win and the \emph{slowest} loss, and so mate scores
  remain comparable across depths. The TT adds/removes the ply on store/probe so a
  mate seen through a transposition keeps the right distance.
\item[Q. Is null-move pruning safe? What about zugzwang?] NMP assumes passing is
  worse than your best real move; that can fail in zugzwang. We disable NMP in
  check and at PV nodes, require depth $\ge 3$, and never return an
  \emph{unproven} mate from a null search (clamp to $\beta$). On this material-rich
  small board zugzwang is rare, and the guards make the rare case safe.
\item[Q. Can LMR/futility/LMP make you miss the best move?] LMR re-searches at full
  depth whenever a reduced search beats $\alpha$, so a good move is never lost --
  only confirmed-bad moves are searched shallowly. Futility/LMP only skip
  \emph{quiet} moves at \emph{low} depth when the static eval is far below $\alpha$;
  captures, promotions, killers, and in-check nodes are exempt. These are
  heuristics that trade a tiny tactical risk for much greater depth, which on this
  board is a strong net positive.
\item[Q. Why is a cached TT score trustworthy?] Each entry records the bound type
  alpha-beta proved: \emph{exact} values are reused directly; a \emph{lower} bound
  only when it already exceeds $\beta$; an \emph{upper} bound only when $\le\alpha$
  -- exactly the conditions under which the cached search would have cut off
  anyway. A 32-bit key check guards against index collisions.
\item[Q. What is quiescence search for, and why is stand-pat valid?] It resolves
  captures past the depth horizon so we don't evaluate mid-trade. Stand-pat (the
  static eval) is a valid lower bound because the side to move can normally decline
  to capture, so the true value is at least the static score; if that already
  $\ge\beta$ we cut.
\item[Q. How do you avoid losing on time?] Iterative deepening always has a
  completed shallower result, time is checked every 2048 nodes and the search stops
  with a safety margin before the budget, and the protocol layer emits the best
  move found so far on \code{stop}. So a legal move is always ready before the 2\,s
  cutoff.
\item[Q. Why blend NNUE with a hand-crafted eval instead of using NNUE alone?]
  Measured: pure NNUE is weaker here, and the hand-crafted eval alone is weaker
  than the 45\% blend. The net adds outcome-correlated positional knowledge; the
  heuristic adds tactical robustness and is cheap. The blend beats either alone.
\item[Q. Why such a small net ($360\to32\to1$)? Isn't bigger better?] Bigger nets
  fit the data better but the per-leaf eval is rebuilt every node, so width costs
  search depth: $H{=}128$ lost $\approx4$ plies at 1.5\,s and was no stronger.
  Depth dominates eval sophistication on a $6\times5$ board, so $H{=}32$ is optimal.
\item[Q. You used PyTorch -- isn't that a forbidden third-party library?] PyTorch
  is used \emph{offline} to train; it produces a header of plain \code{float}
  arrays. The competing engine links only the C++ standard library. Training tools
  are not part of the submitted engine.
\item[Q. Why encode features from the side-to-move's perspective?] One weight set
  then serves both colours, halving the parameters and doubling effective data.
  Black's start is exactly White's rotated $180^\circ$, so a single rotation makes
  the mapping exact.
\item[Q. How is the NNUE trained ``from scratch''?] Self-play games (C++ generator
  using our own search) label positions with their game result (and teacher score);
  a small MLP is trained to predict that. No external datasets or pre-trained
  weights.
\item[Q. What stops infinite loops / repetitions?] A Zobrist-keyed history map;
  the third occurrence of a position returns a draw, both implementing the 3-fold
  rule and terminating cyclic search lines.
\item[Q. Why bitboards on a 5-wide board?] All 30 squares fit in a
  \code{uint32\_t}; a single \code{\_\_builtin\_ctz} bit-scan plus precomputed
  leaper masks replaces nested array scans, and slider rays are short. It is both
  faster and simpler than the naive generator (which we keep for reference).
\item[Q. What is the single most important factor in your engine's strength?]
  Search depth, enabled by good move ordering (TT move + MVV-LVA + killers/history)
  and safe pruning. The eval and NNUE refine the leaves, but reaching one more ply
  reliably beats a smarter but slower evaluation here.
\end{description}

% ==================================================================
\section{Key points to remember for the demo}
% ==================================================================
\begin{enumerate}[nosep]
  \item \textbf{Search}: negamax + alpha-beta + PVS + quiescence, plus TT,
        killer/counter/history ordering, IID, aspiration windows, NMP, LMR,
        reverse-futility/futility/late-move pruning, mate-distance pruning, and
        check extensions. Be ready to explain \emph{why} each prunes safely
        (Section~\ref{sec:qa}).
  \item \textbf{Evaluation}: material + PST + king tropism + pawn structure +
        passed pawns + mobility, \emph{blended} with a from-scratch NNUE at 45\%.
  \item \textbf{NNUE}: 360-feature side-to-move-relative encoding ($180^\circ$
        flip for Black), $360\!\to\!32\!\to\!1$; trained on $\approx 926\text{k}$
        positions from strong-search self-play (C++ generator) with a PyTorch GPU
        trainer, on a blended WDL\,+\,teacher-score target; weights embedded in a
        header; sparse per-leaf inference.
  \item \textbf{King-capture model}: no checkmate/legality logic needed; the engine
        cannot output an illegal move; ``win'' = a move takes the king.
  \item \textbf{Headline insight}: \emph{on a small board, depth beats eval
        sophistication}; SOTA chess NNUE architecture and its quantization trick do
        \emph{not} transfer (measured, not assumed).
  \item \textbf{Second insight}: \emph{lower training loss $\ne$ stronger play} --
        shown three times (wider NNUE, dual-perspective NNUE, texel tuning).
  \item \textbf{Methodology}: use random openings to measure strength; a speedup
        (node pool) must be a provable non-regression (identical node counts)
        before shipping.
\end{enumerate}

\paragraph{Build.}
\begin{lstlisting}[language=bash]
g++ --std=c++2a -Wall -Wextra -Wpedantic -g -O3 -march=native -static \
  -Isrc/games/minichess -Isrc/state -Isrc \
  -o build/minichess-ubgi \
  src/games/minichess/state.cpp src/policy/minimax.cpp \
  src/policy/random.cpp src/policy/submission.cpp src/ubgi/ubgi.cpp
\end{lstlisting}
NNUE is on by default and self-contained; \code{-DNO\_NNUE} disables it and
\code{-DNNUE\_BLEND=$n$} changes the mix. Retrain with: build
\code{tools/nnue\_gen.cpp}, run several shards (\code{nnue\_gen out.bin games depth
seed}), then \code{python tools/train\_nnue.py --hidden 32}.

\end{document}
